---
title: "Data wrangling"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
    theme: paper
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```

```{r xaringanExtra-clipboard, echo=FALSE}
xaringanExtra::use_clipboard()
```

Download the script for this session at the link below (Click Link --> Right Click on Page --> Save Page As):

* [Session 3 and 4: data wrangling R script](https://raw.githubusercontent.com/jdmwhite/R_Workshop/main/scripts/3_Data_Wrangling.R)

Data wrangling involves cleaning, (re)structuring and transforming the data. This allows us to analyse and visualise data more easily.

# Part 1: Data cleaning

In this section we will look at turning messy data into tidy data. Tidy data allows us easily (re)structure it, should we need to. Cleaning data involves removing unnecessary spaces, periods and duplicate observations in the data. Clean data make it easy for us to select columns and rows in our data set.

## Install packages and load libraries

We will start by installing and loading the packages that we will need to clean our data and for the rest of the data wrangling session. *You might already have these packages. If you do, you can skip this step. We will be using the janitor package as well as the dplyr, tidyr and readxl packages from the tidyverse collection of packages. 

```{r eval=FALSE}
#### Install packages ----
install.packages('tidyverse') # dplyr, tidyr, readxl  
install.packages('janitor') # cleaning names
```

Next we need to load our packages (libraries) onto our work spaces. Note that we have to load 'readxl' separately because it is not a core tidyverse package like dplyr and tidyr. 

```{r, message = FALSE, warning=FALSE}
#### Load libraries ----
library(readxl)
library(tidyverse)
library(janitor)
library(dplyr)
```

## Import data

Our data that need cleaning will be loaded next. Data can be imported from .txt, .csv or Excel (to name a few common file types) into R. For this workshop we will be importing an excel file. I added another line to show one can import a .csv file too.

```{r, message = FALSE, warning=FALSE}
#### Loading the data ----
photographer_names <- read_excel("data/raw/photographer_names.xlsx")
#photographer_names <- read.csv("path/filename.csv)
```
## Inspect data

Before we decide how to clean the data, we need to have a clear view of the data and know how they are structured. We want to find out information about the rows(observations) and columns(variables). This will give us an idea of how we can clean, manipulate/transform and (re)structure the data before visualisation and analysis. 

```{r, message = FALSE, warning=FALSE}
#### Inspecting the data
View(photographer_names)
names(photographer_names) # gives you the names of the columns 
str(photographer_names) # gives you the structure of the data
head(photographer_names) # gives you the first 6 rows of the data
tail(photographer_names) # gives you the last 6 rows of the data
```

## Prepare data

Data preparation is a step that you might not always need if your data have no gaps, extra letter, symbols or other attibutes you might not need. It also allows you to see duplicate information. Preparing the data allows you to clean up the messy parts of your data that may make it hard for you to transform your data.We will start by cleaning the names of the columns, renaming columns that have ambiguous names. Then we will move to changing the column class. Data can be in different classes/formats such as numeric, character or date format.   

```{r, message = FALSE, warning=FALSE}
#### Clean and rename data (columns)
photographers <- clean_names(photographer_names) # the new names are unique
photographers <- rename(photographers, initial_date = date_4)
photographers <- rename(photographers, repeat_date = date_6)
names(photographers) # view the new names of the columns

#### changing column format/class
class(photographers$initial_date) # view the class of the column
class(photographers$repeat_date)

photographers %>% mutate(initial_date = case_when(
  initial_date == "1986/7" ~ 
    as.character(as.Date('86/01/01', format = "%y/%m/%d")),
  initial_date != "1986/7" ~ 
    as.character(as.Date(as.numeric(initial_date), origin="1899-12-30"))
)) -> photographers

View(photographers)

#### remove or add rows
rows_delete (photographers, tibble(site_number = 1100)) #you can match the delete function to any specific row
rows_insert (photographers, tibble (site_number = 9)) #The new row is added.
#### remove or add columns
select(photographers,-initial_date) #removes column
photographers <- mutate (photographers, location = " ") #adds a new column

```

# Part 2: Data transformation

Data transformation involves changing the structure of data.This involves adding new columns with computed data or joining a new column/row to the dataframe. We did some transformations above. Here are a few more.

```{r, message = FALSE, warning=FALSE}
#### Clean and structure data
pull(photographers, latitude)
select(photographers, site_location, altitude_m)

```


