---
title: "Data analysis"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 4
    theme: paper
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```

```{r xaringanExtra-clipboard, echo=FALSE}
xaringanExtra::use_clipboard()
```
# Basic data analysis

The most basic computing one can do on R is arithmetic
```{r}
1+3

```
```{r}
1+(4*2)

```
```{r}
1+(4^2) ##^: exponent

```
Trigonometric functions
```{r}
sin(30)

```
```{r}
log10 (10)
log10 (20)

```
Comparisons 

```{r}
2 == 2

```
```{r}
2 == 4

```
```{r}
1 < 2

```


# Data analysis using a dataframe

We will continue to work with the penguins data from the data visualisation section. These data have already been cleaned and visualised. Therefore, we will only focus on the analysis. 

load the processed data

```{r}
penguins <- palmerpenguins::penguins
str(penguins)
View(penguins)
```

## Linear models
```{r}
oneway <- lm(bill_length_mm ~ island, data = penguins)
oneway
```
## dplyr, broom & stargazer