---
title: "Data visualisation"
output:
  html_document:    
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
    theme: paper
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```

# Part 1: Building a ggplot()

In Part 1, we will work on building one good, publication ready figure.

Install the required packages. Each package has a comment next to it with a brief reason why we're using it.

```{r eval=FALSE}
#### Install packages ----
install.packages('tidyverse') # dplyr and ggplot2
install.packages('janitor') # cleaning names
install.packages('lubridate') # extracting dates
install.packages('stringr') # extracting strings
install.packages('patchwork') # combining plots
install.packages('palmerpenguins') # the penguin data!
```

Load your libraries:

```{r, message = FALSE, warning=FALSE}
#### Load libraries ----
library(tidyverse)
library(janitor)
library(lubridate)
library(stringr)
library(patchwork)
library(palmerpenguins)
```

## Explore the data

The penguin package has a great collection of data, in both raw and processed formats. Let's take a look at it:

```{r}
#### Load data ----
data(package = 'palmerpenguins')
```

First let's explore the raw data:

```{r}
#### Explore data ----
penguins_raw <- palmerpenguins::penguins_raw
str(penguins_raw)
```

Second, take a look at the processed data:

```{r}
penguins <- palmerpenguins::penguins
str(penguins)
```

## Wrangle the data

We will start working on the raw data and process it so that it looks just like the clean.

Start by using the `janitor::clean_names()` function to remove any spaces in the column names.

```{r}
#### Wrangle data ----
# clean the column names
penguins_clean_1 <- clean_names(penguins_raw)
names(penguins_clean_1)
```

We then select the columns that we'd like to work with. In the `dplyr` package, we have previously used the `rename()` function. However, we can use `select()` to both select the columns of interest and rename the columns.

Here we rename the columns with *culmen* to *bill* and change the *date* column to *year*.

```{r}
# select columns to work on & clean names
names(penguins_clean_1)
penguins_clean_1 %>% select(species, island, bill_length_mm = culmen_length_mm, bill_depth_mm = culmen_depth_mm, flipper_length_mm, body_mass_g, sex, year = date_egg) -> penguins_clean_2
```

The date column was previously stored as string. Let's extract only the year using the `lubridate::year()` function.

```{r}
# select only year from dates
penguins_clean_2$year <- year(penguins_clean_2$year)
```

We're getting close to the processed data. Take another look at the data structures:

```{r}
# check that they are the same
str(penguins)
str(penguins_clean_2)
```

To clean the species column, we need to select only the first word from the string. To do this, we use the `stringr::word()` function and specify the 1st word. We also wrap this in the `as.factor()` function to convert the column from character to factor.

```{r}
# select only first name from species list and convert to factor
penguins_clean_2$species <- as.factor(word(penguins_clean_2$species, 1))
```

Lastly, let's convert the *sex* column values to lower case using the base R function `tolower()`. Again, convert this to a factor.

```{r}
# change sex to lower case
penguins_clean_2$sex <- as.factor(tolower((penguins_clean_2$sex)))
```

And then a final check. Save the cleaned data over the processed data that we loaded in earlier.

```{r}
# final check
str(penguins_clean_2)
penguins <- penguins_clean_2
```

## Data visualisation

Now that our data is cleaned, we can begin plotting. Let's first look at the broad relationships between all of the variables in our data. We can simply use the base R `plot()` to explore all of the correlations in our data.

```{r}
#### Data visualisation ----
# correlation plot with all variables
plot(penguins)
```

From this plot, we can see that **body mass** and **bill length** look strongly correlated. Let's use these two variables to build a plot.

In this workshop, we will be using the `ggplot2` package for *most* of our data visualisation. There are several different R and non-R options for plotting data, so why chose `ggplot2`?

-   It is like building lego blocks (each geom is a layer)
-   The syntax is *relatively* simple
-   Easy use of themes & colour palettes
-   Wide adoption (so easy to find similar queries online)
-   Facets! (easily produce several subplots based on a factor)
-   Combine plots seamlessly (there are several packages for this, but we'll explore `patchwork`. Also look up `cowplot` or the `egg` package)
-   `gganimate()`: easily turn your plot into an animation (we don't explore this function here, but take a look at the [great examples!](https://gganimate.com/))
-   `ggplot2` is a part of the `tidyverse` group of R packages. This means that it is well integrated with some of the other packages we have learnt so far, like `dplyr`.

Ok, those are a few reasons. However, always do what's easiest for youm whether it's using base R functions, excel charts, plotnine/seaborn/matplotlib in Python or whatever else. `ggplot2` is just one of these options and the most important thing is to get the job done.

The primary function we will use is `ggplot()` (don't forget to remove the 2!). The core template used in `ggplot2` graphics requires the `data.frame`, the variables of interest and the type of plot you want. This is the template:

`ggplot(data = DATA, mapping = aes(VARIABLES OF INTEREST)) + geom_()`

`ggplot2` works similarly to `dplyr`'s `%>%` (pipe) function, but instead we use a `+` to add on additional functions. Find the `ggplot2` cheatsheet [HERE](https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf).

Let's get started by first selecting the dataset we want to work with:

```{r}
# chose the data
ggplot(data = penguins)
```

This doesn't show us anything, because we have not yet chosen our variables. We then chose our variables of interest and place them inside our mapping aesthetics. Here we want `body_mass_g` on the x-axis and `bill_length_g` on the y-axis.

```{r}
# add in mapping aesthetics
ggplot(data = penguins, mapping = aes(x = body_mass_g, y = bill_length_mm))
```

### Basic plot

Our axis are set up! Let's now fill the plot by choosing a `geom_()` function, short for **geometry**. We will use `geom_point()`, which plots a scatterplot.

```{r}
# chose a geometry to plot
ggplot(data = penguins, mapping = aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point()
```

There's your first `ggplot()`! However, there are several options to start customising it to our liking, so let's explore those. Another option, is the place the `data =` and `mapping =` arguments within the `geom_()`.

```{r}
# add the data and mapping arguments to the geom
ggplot() +
  geom_point(data = penguins, mapping = aes(x = body_mass_g, y = bill_length_mm))
```

We can also do a mix of the two, as shown below. Note that if our arguments are in the correct place, we can also remove the calls to `data =` and `mapping =`. This helps to shorten and clean up our code:

```{r}
# keep the data in ggplot() and mapping aesthetics in the geom_
ggplot(penguins) +
  geom_point(aes(body_mass_g, bill_length_mm))
```

### Add in colour

Let's now add a 3rd variable `species` as the **colour** using `col =`.

```{r}
# add in a 3rd variable as colour
ggplot(penguins) +
  geom_point(aes(body_mass_g, bill_length_mm, col = species))
```

### Add in shape

Notice that `ggplot()` automatically adds in a legend! We can double up, and also change the shape of each point based on the `species` using `shape =`.

```{r}
# add in a 3rd variable as colour and shape
ggplot(penguins) +
  geom_point(aes(body_mass_g, bill_length_mm, col = species, shape = species))
```

### Change opacity

Note that these two aesthetics are placed within the same mapping aesthetics as the x and the y variables. This is **very important**. If we placed them outside of the `aes()` call, it would not work. Anything outside of the `aes()` call will be applied to all points! See the example below on changing the opacity using `alpha =`. Changing the opacity helps to show points, which may be overlapping with others.

```{r}
# change the opacity of the points
ggplot(penguins) +
  geom_point(aes(body_mass_g, bill_length_mm, col = species, shape = species), alpha = 0.8)
```

### Change the colours

Let's further customise our plot by changing the default colours to those we would like to use. To select colours, we can specify colour names (e.g. 'blue') or use a hex colour code. See a list of the colour names found in R [HERE](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf). There are several useful tools to select hex colour codes, [HERE](https://htmlcolorcodes.com/) is one example.

To change the colours we use the `scale_` functions. Importantly, `scale_` can be used for many purposes, including changing axes limits or breaks or transforming your axes. We will use it to change the shape and colours of our plots.

If you are using the `col =` in your `geom_`, then we need to use `scale_colour_` in our `scale_` function. The same applies to `shape =` and `fill =` (which we will use later...). The last part of our `scale_colour` function is to chose whether it's a *discrete* or *continuous* variable. In our case, we have a discrete variable, but instead of using `scale_colour_discrete()`, we will use `scale_colour_manual()`, so that we can manually select our colour values.

We also use the `labs()` function to change the legend title. If we only do this for colour, then the shape aesthetic will still use the original title and the `ggplot()` will plot two separate legends. We don't want this, so we specify both titles, which then combines our legends together:

```{r}
# change the colour of the points and the legend name (for both scales)
# use c('darkorange', 'purple', 'cyan4')
ggplot(penguins) +
  geom_point(aes(body_mass_g, bill_length_mm, col = species, shape = species), alpha = 0.8) +
  scale_colour_manual(values = c('darkorange', 'purple', 'cyan4')) +
  labs(colour = 'Penguin species', shape = 'Penguin species')
```

### Change the labels

In our `labs()` function, we can also add in our x- and y-axes labels:

```{r}
# change the labels of the x and y axes
ggplot(penguins) +
  geom_point(aes(body_mass_g, bill_length_mm, col = species, shape = species), alpha = 0.8) +
  scale_colour_manual(values = c('darkorange', 'purple', 'cyan4')) +
  labs(x = 'Body mass (g)', y = 'Bill length (mm)',
       colour = 'Penguin species', shape = 'Penguin species') 
```

### Change the axes

As mentioned before, `scale_` can also be used to change our axes limits and breaks. If you have a discrete x-axis, you would use `scale_x_discrete()`, but in this case both of our axes are continuous, so we use `scale_x_continuous()`. Here is an example, which changes our limits using lower- and upper-bounds (i.e. the zoom-level) and our breaks (i.e. what tick marks would you like).

```{r}
# change the limits and breaks of the x and y axes
ggplot(penguins) +
  geom_point(aes(body_mass_g, bill_length_mm, col = species, shape = species), alpha = 0.8) +
  scale_colour_manual(values = c('darkorange', 'purple', 'cyan4')) +
  labs(x = 'Body mass (g)', y = 'Bill length (mm)',
       colour = 'Penguin species', shape = 'Penguin species') +
  scale_x_continuous(limits = c(2500, 6500), breaks = seq(2500,6500,1000)) +
  scale_y_continuous(limits = c(30,60), breaks = c(30,40,50,60))
```

### Themes

One of my favourite features of `ggplot2`, is the easily editable themes. There are several built-in themes, which can quickly convert your plots into clean designs. All themes can be use with the basic call to `theme_`. Here will use the minimal theme `theme_minimal()`.

```{r}
# change the default theme
ggplot(penguins) +
  geom_point(aes(body_mass_g, bill_length_mm, col = species, shape = species), alpha = 0.8) +
  scale_colour_manual(values = c('darkorange', 'purple', 'cyan4')) +
  labs(x = 'Body mass (g)', y = 'Bill length (mm)',
       colour = 'Penguin species', shape = 'Penguin species') +
  theme_minimal() 
```

### Legend position

We can also edit where we put the legend. We change this (as well as many other customisable options) in the `theme()` function. We have a few default options for the legend, which include `"top"`,`"bottom"`,`"left"` and `"right"`. However, we can also specify the position *inside* the plot using a values for the x- and y-axes, respectively. These values fall between 0 and 1.

```{r}
# change the legend position
# ?theme
ggplot(penguins) +
  geom_point(aes(body_mass_g, bill_length_mm, col = species, shape = species), alpha = 0.8) +
  scale_colour_manual(values = c('darkorange', 'purple', 'cyan4')) +
  labs(x = 'Body mass (g)', y = 'Bill length (mm)',
       colour = 'Penguin species', shape = 'Penguin species') +
  theme_minimal() +
  theme(legend.position = c(0.8,0.2))
```

In my books, this is a complete and ready to publish figure! We could save it to file at this point. But before we do that, let's explore a few more `ggplot2` fundamentals.

### Facets

One of the powerful features of `ggplot2` is the ability to facet a plot by one of your discrete variables (i.e. factors). This means we can use the same plot recipe as used above, but adding just one more line of code, we can split it up into several subplots using the `facet_grid()` or `facet_wrap()` functions. Here, we split our plot into two subplots by `sex` and stack them next to each other using `cols = vars(sex)`. If we want them on top of one another, we would use `rows =`.

We can also tweak the theme by using `theme_bw()`, which works nicely with facets.

```{r}
#### Facets ----
ggplot(penguins %>% drop_na()) +
  geom_point(aes(body_mass_g, bill_length_mm, col = species, shape = species), alpha = 0.8) +
  scale_colour_manual(values = c('darkorange', 'purple', 'cyan4')) +
  labs(x = 'Body mass (g)', y = 'Bill length (mm)',
       colour = 'Penguin species', shape = 'Penguin species') +
  theme_bw() +
  facet_grid(cols = vars(sex))
```

### Combine plots

So facets can help us visualise several subplots based on a factor, but what if we want to plot two completely separate plots next to one another?

There are several packages which could help us here, but let's explore a more recent and extremely useful one called `patchwork`. `patchwork` uses a basic syntax to combine plots together.

-   To add plots as columns use `+`
-   To add plots as rows use `/`

There is a huge variety of ways to patch plots together, so take a look at the [patchwork website](https://patchwork.data-imaginist.com/index.html).

We will create a new plot (body mass versus flipper length) and save each plot to an object (`plot1` and `plot2`).

```{r}
#### Patchwork ----
plot1 <- ggplot(penguins) +
  geom_point(aes(body_mass_g, bill_length_mm, col = species, shape = species), alpha = 0.8) +
  scale_colour_manual(values = c('darkorange', 'purple', 'cyan4')) +
  labs(x = 'Body mass (g)', y = 'Bill length (mm)',
       colour = 'Penguin species', shape = 'Penguin species') 

plot2 <- ggplot(penguins) +
  geom_point(aes(body_mass_g, flipper_length_mm, col = species, shape = species), alpha = 0.8) +
  scale_colour_manual(values = c('darkorange', 'purple', 'cyan4')) +
  labs(x = 'Body mass (g)', y = 'Flipper length (mm)',
       colour = 'Penguin species', shape = 'Penguin species') 

# combine plots (cols and rows)
plot1 + plot2 # cols
plot1 / plot2 # rows
```

We can then save our two patched together plots as a new object `combined_plot`.

```{r}
combined_plot <- plot1 + plot2 
combined_plot
```

`patchwork` also offers some really valuable annotation and layout control options. For example, we can add plot annotations to give them clear labels:

```{r}
# add annotation
combined_plot + plot_annotation(tag_levels = 'a', tag_suffix = ')')
```

Seeing as our plots have the same legend, we can also combine these together:

```{r}
# control the guides
combined_plot + plot_layout(guides = 'collect')
```

A really useful helper, is to apply a theme to all of the plots. Note that we are now using `&` instead of `+`. This is because if we were to use `+` it would only change the theme for our last plot. `&` makes sure this applies to both/all plots:

```{r}
# add theme style to all plots
combined_plot  &
  theme_minimal() &
  theme(legend.position = 'bottom')
```

And now we can finally bring all of the patchwork edits together:

```{r}
# bring it all together
combined_plot + 
  plot_annotation(tag_levels = 'a', tag_suffix = ')') + 
  plot_layout(guides = 'collect') &
  theme_minimal() &
  theme(legend.position = 'bottom')
```

## Save your ggplot

Now that we have our publication ready plot(s), we can save our ggplot!

This requires the `ggsave()` function. `ggsave()` will automatically save the last plot we have plotted in our plot window. This means we just need to specify the file directory and the device (i.e. what type of output would you like). However, there are some valuable additions to this function. For example, if we use `.png`, we may want to specify the background colour using `bg =`. Importantly, we can also choose our plot dimensions. This is crucial if you creating plots for a publication, where many of them have size and format requirements! Change the units to mm using `units = 'mm'` and then chose your width and height.

```{r, eval = FALSE}
#### Save your plot ----
# using ggsave()
ggsave("output/figs/peng_mass_bill_flipper_length.png", device = 'png', bg = 'white', width = 120, height = 80, units = 'mm', dpi = 320)
```

# Part 2: Different geometries

In Part 2, we will explore the range of different plots available in `ggplot2`. We have already looked at a type of correlation plot using `geom_point()` (i.e. scatterplots). In this section, we will explore a range of distribution plots (boxplots, histograms and density plots) and ranking plots (barplots and lollipop plots). There are many more ways to plot your data and the [r-graph-gallery](https://r-graph-gallery.com/index.html) or following the [#TidyTuesday challenge on Twitter](https://twitter.com/search?q=%23TidyTuesday&src=typeahead_click) are good places to look for inspiration.

Again, we start by installing & loading our packages:

```{r, eval=FALSE}
#### Install packages ----
install.packages('tidyverse') # for dplyr and ggplot2
install.packages('palmerpenguins') # our penguin data!
install.packages('ggrepel') # for repelled text labels
install.packages('ggridges') # for interesting density plots
```

```{r}
#### Load libraries ----
library(tidyverse)
library(palmerpenguins)
library(ggrepel)
library(ggridges)
```

We will use the pre-processed penguin data this time:

```{r}
#### Explore data ----
penguins <- palmerpenguins::penguins
str(penguins)
```

Using the same plot as before, let's now add linear trendlines to each of our species. Notice that we have used the function `drop_na()` to remove any `NA` values in our penguin data. To add our trendlines, we use `geom_smooth()`. The default `method =` for `geom_smooth()` is `loess`, which provides a smoothed loess curve to each species. To change this, we select `method = 'lm'`:

```{r}
#### Line charts ----
ggplot(data = drop_na(penguins), aes(body_mass_g, bill_length_mm, col = species, shape = species)) +
  geom_point(alpha = 0.8) +
  scale_colour_manual(values = c('darkorange', 'purple', 'cyan4'),name = 'Penguin species') +
  scale_shape(name = 'Penguin species') +
  geom_smooth(method = 'lm') +
  labs(x = 'Body mass (g)', y = 'Bill length (mm)') +
  theme_bw() 
```

A plot type you'll all be familiar with is a barplot. A barplot requires summarised data. So let's do some quick data manipulation. First, `group_by(species)` and then summarise **flipper_length_mm** by `mean()` and `sd()`. Notice we use `na.rm = TRUE` to remove the 2 rows with missing data.

We can then `%>%` (pipe) our summarised dataset directly into a `ggplot()`! To produce a barplot, we use `geom_col()` and we add our errorbars (SD) by providing an `x` value, as well as a `ymax` and `ymin` value. We need to calculate these on the fly, so we take the `flipper_mean` and then `+` or `-` the `flipper_sd` values, respectively. It's worth adjusting the `width =` of the errorbars to produce a more pleasing appearance:

```{r}
#### Barplot with SD ----
penguins %>% 
  group_by(species) %>% 
  summarise(flipper_mean = mean(flipper_length_mm, na.rm = T),
            flipper_sd = sd(flipper_length_mm, na.rm = T)) %>% 
  ggplot() +
  geom_col(aes(x = species, y = flipper_mean),
           fill = c('darkorange','purple','cyan4')) +
  geom_errorbar(aes(x = species, ymax = flipper_mean + flipper_sd, ymin = flipper_mean - flipper_sd), width = 0.5) +
  labs(x = "Penguin species", y = 'Flipper length (mm)') +
  theme_minimal()
```

Barplots aren't great. They tend to hide the underlying distribution of the data, which is what most people are actually interested in. The only time you should use them is when you only have a single value for each variable (i.e. no variability). But even then, there are far more interesting options ([HERE](https://towardsdatascience.com/anything-but-bars-the-10-best-alternatives-to-bar-graphs-fecb2aaee53a) and [HERE](https://www.r-bloggers.com/2017/06/bar-plots-and-modern-alternatives/)). Read more about why to avoid them [HERE](https://paulvanderlaken.com/2018/12/17/avoid-bar-plots-for-continuous-data-do-this-instead/).

We do have variation in our data, so let's look at better alternatives. First, we have the classic box-and-whisker plots. These start to give us a better idea of our distribution:

```{r}
#### Boxplot ----
ggplot(penguins) +
  geom_boxplot(aes(x = species, y = flipper_length_mm),
               col = c('darkorange','purple','cyan4')) +
  labs(x = "Penguin species", y = 'Flipper length (mm)') +
  theme_minimal() 
```

We're still not quite there. Histograms are an excellent tool to show us the distribution of our data clearly. In the `geom_histogram()` geom, select `position = 'identity` and `alpha = 0.6` to avoid stacked frequency bars and to make overlaps more obvious:

```{r}
#### Histograms ----
ggplot(penguins) +
  geom_histogram(aes(x = flipper_length_mm, fill = species), position = 'identity', alpha = 0.6) +
  scale_fill_manual(values = c('darkorange','purple','cyan4')) +
  labs(x = 'Flipper length (mm)', y = 'Frequency') +
  theme_minimal()
```

We could also use density plots, which essentially show a smoothed histogram output:

```{r}
#### Density ----
ggplot(penguins) +
  geom_density(aes(x = flipper_length_mm, fill = species), col = NA, alpha = 0.6) +
  scale_fill_manual(values = c('darkorange','purple','cyan4')) +
  labs(x = 'Flipper length (mm)', y = 'Density') +
  theme_minimal()
```

There are several R packages which work together with `ggplot2`. In this example, we use the `geom_density_ridges()` function from the `ggridges` package. This places each density ridge on its own line:

```{r}
#### Ridges ----
ggplot(penguins) +
  geom_density_ridges(aes(x = flipper_length_mm, y = species, fill = species), col = NA, alpha = 0.6) +
  scale_fill_manual(values = c('darkorange','purple','cyan4')) +
  labs(x = 'Flipper length (mm)', y = 'Density') +
  theme_minimal()
```

```{r}
#### Lollipop chart ----
flipper_mean_all <- mean(penguins$flipper_length_mm, na.rm = T)

peng_summary <- penguins %>% 
  group_by(species) %>% 
  summarise(flipper_mean = mean(flipper_length_mm, na.rm = T),
            flipper_sd = sd(flipper_length_mm, na.rm = T)) 

ggplot() +
   geom_segment(data = peng_summary, aes(x = species, xend = species,
                    y = flipper_mean_all, yend = flipper_mean, col = species)) +
  geom_point(data = peng_summary, aes(x = species, y = flipper_mean, col = species)) +
  scale_color_manual(values = c('darkorange','purple','cyan4'), name = 'Penguin species') +
  geom_hline(aes(yintercept = flipper_mean_all), lty = 2) +
  labs(x = '', y = 'Mean flipper length (mm)') +
  coord_flip() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = 'top') +
  geom_text_repel(aes(x = 1.5, y = flipper_mean_all, label = 'overall mean'), nudge_x = 0.5, nudge_y = 7.5, segment.curvature = -0.1, size = 3, segment.size = 0.2)
```
