---
title: "Github and Rmarkdown"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
    theme: paper
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```

```{r xaringanExtra-clipboard, echo=FALSE}
xaringanExtra::use_clipboard()
```

## Background

In this topic, we will cover:

* What is Github?
* Why should you use Github?
* What is Rmarkdown?
* (and you guessed it) Why should you use Rmarkdown?

Neither of these platforms/tools are essential to moving along the R experience ladder. The reason I decided to include them in a beginners R workshop is because they allow for some extremely valuable workflows related to sharing and collaboration. So, I think at least a little bit of exposure to these is useful early on in your RStudio journey.

## What is Github?

## Why should you use Github?

## What is Rmarkdown?

## Why should you use Rmarkdown?
