---
title: "R Markdown and Github"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
    theme: paper
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```

```{r xaringanExtra-clipboard, echo=FALSE}
xaringanExtra::use_clipboard()
```

## Background

In this topic, we will cover:

* What is R Markdown?
* Why should you use R Markdown?
* R Markdown resources
* What is Github?
* Why should you use Github?
* Github-RStudio resources

Neither of these platforms/tools are essential to moving along the R experience ladder. The reason I decided to include them in a beginners R workshop is because they allow for some extremely valuable workflows related to sharing and collaboration. So, I think at least a little bit of exposure to these is useful early on in your RStudio journey.

## What is R Markdown?

> "R Markdown provides an authoring framework for data science"
`r tufte::quote_footer('RStudio')`

Using R Markdown, you are able to run code, share its output and provide commentary in the same document. This allows you to produce high quality R reports that you can share with an audience. For example, this entire R Workshop webpage was built using R Markdown. You are welcome to view and download the [original R Markdown files](https://github.com/jdmwhite/R_Workshop) to see how it was put together. 

It is a format of markdown, which is a plain text format (as opposed to a rich text format like MS Word).

The plain text format can make the barrier to entry for R Markdown a little bit tricky. But just like R scripts there are lots of examples out there and once you've tested it out you will get the hang of it quite quickly.

There are 3 major components to an R Markdown file:

1. a YAML header denoted by the `---`s (this is optional)
2. R code chunks denoted by ` ``` `s
3. Plain text combined with text formatting

These three components are all you need to produce a high quality report in R. Here is how they are 'coded' in an R markdown file:

1. YAML header
```yaml
---
title: "Title here"
author: "Author here"
output:
    pdf_document
---
```

2. R code chunk

````markdown
`r ''````{r}
1 + 1
```
````

```{r, echo = FALSE}
1 + 1
```

````markdown
`r ''````{r}
plot(x = seq(1:10), y = seq(1:10))
```
````
```{r, echo = FALSE}
plot(x = seq(1:10), y = seq(1:10))
```

3. Plain text with formatting

```markdown
# First-level header

## Second-level header

### Third-level header

*italics*

**bold**
```

## Why should you use R Markdown?

* Sharing work with collaborators
* Producing high quality reports that integrates R outputs with text
* For educators: coursework hand-ins

## R Markdown resources

* [https://bookdown.org/yihui/rmarkdown/](https://bookdown.org/yihui/rmarkdown/)
* [R Markdown Cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/rmarkdown-2.0.pdf)
* [R Markdown Reference Guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)

## What is Github?

## Why should you use Github?

## Github-RStudio resources

* https://happygitwithr.com/
* https://resources.github.com/github-and-rstudio/
