---
title: "R Markdown and Github"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
    theme: paper
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```

```{r xaringanExtra-clipboard, echo=FALSE}
xaringanExtra::use_clipboard()
```

## Background

In this topic, we will cover:

* What is R Markdown?
* Why should you use R Markdown?
* R Markdown resources
* What is Github?
* Why should you use Github?
* Github-RStudio resources

Neither of these platforms/tools are essential to moving along the R experience ladder. The reason I decided to include them in a beginners R workshop is because they allow for some extremely valuable workflows related to sharing and collaboration. So, I think at least a little bit of exposure to these is useful early on in your RStudio journey.

## What is R Markdown?

> "Rmarkdown is file format for making dynamic documents with R"
`r tufte::quote_footer('--- RStudio')`

It is a format of markdown, which is a plain text format (as opposed to a rich text format like MS Word)

## Why should you use R Markdown?

## R Markdown resources

* [https://bookdown.org/yihui/rmarkdown/](https://bookdown.org/yihui/rmarkdown/)

## What is Github?

## Why should you use Github?

## Github-RStudio resources

* https://happygitwithr.com/
* https://resources.github.com/github-and-rstudio/
